Fast and adaptive cointegration based model for forecasting high frequency financial time series

### 2017-01-16 articles-conferences/iscef2016/body.tex
intro re estructurado

VECM parameters are obtained using the ordinary least squares (OLS) method,
developed by \cite{golub1980}. Since OLS involves many calculations, the
parameter estimation is computationally expensive when the number of lagged
values and data increases. Moreover, obtaining cointegration vectors is also an
expensive routine. This is a main limitation to use VECM with stream data with
high frequency. \cite{chen2003} addresses the advantage of distributed
processing over conventional rolling window processing.

Cointegration relationship can be found in low and high frequency data of two
or more assets. While cointegration in low frequency data is motivated by a
long-run equilibrium relationship between economic forces, cointegration in
high frequency data has its foundation in statistical arbitrage which is very
helpful to detect mean-reverting trades. Cointegration at different data
frequencies may differ. \cite{rittler2012} shows how cointegration appears when
the data frequency is increased. Our aims were to study if a parallel
version of VECM can be used with high frequency stream data.

Our approach was to determine adaptively the number of observations and lags of
VECM which maximise cointegration relations in the past.  This search is
computationally expensive because the Johansen method, see \cite{johansen1995},
is required to find cointegration vectors, which is of order $O(n^3)$.
Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), is
to parallelise this search in order to get new parameters before new data
arrives.  Model effectiveness is focused on out-of-sample forecast rather than
in-sample fitting. This criterion allows AVECM prediction capability to be
expressed rather than just explaining data history.  The forecast capability of
our method was measured using MSE and the Theil's $U$-statistic, see
\cite{theil1966}, widely used in economic forecast. Tests were run using four
currency rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British
Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to
Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

@@

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{Fig1}
  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.9\textwidth]{Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{Fig3}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

>>>>

Cointegration relationship can be found in low and high frequency data of two
or more assets. While cointegration in low frequency data is motivated by a
long-run equilibrium relationship between economic forces, cointegration in
high frequency data has its foundation in statistical arbitrage which is very
helpful to detect mean-reverting trades. Cointegration at different data
frequencies may differ. \cite{rittler2012} shows how cointegration appears when
the data frequency is increased.

The use of VECM with high frequency data is mainly limited by computational 
expensive routines. Firstly, VECM parameters are obtained using the ordinary least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves many calculations, the
parameter estimation is computationally expensive when the number of lagged
values and data increases. Secondly, obtaining cointegration vectors is also an
expensive routine because the Johansen method, see \cite{johansen1995}, is required, which is of order $O(n^3)$. \cite{chen2003} addresses the advantage of distributed
processing over conventional rolling window processing. Therefore, our aims were to study if a parallel version of VECM can be used with high frequency stream data.

Our approach was to determine adaptively the number of observations and lags of
VECM which maximise cointegration relations in the past in a distributed environment. 
We called Adaptively Vector Error Correction (AVECM) to our proposal. AVECM
parallelise this search of parameters in order to update them before new data
arrives. Model effectiveness is focused on out-of-sample forecast rather than
in-sample fitting. This criterion allows AVECM prediction capability to be
expressed rather than just explaining data history.  The forecast capability of
our method was measured using MSE and the Theil's $U$-statistic, see
\cite{theil1966}, widely used in economic forecast. Tests were run using four
currency rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British
Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to
Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

@@

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  %\includegraphics[width=0.9\textwidth]{Fig1}
   \includegraphics[width=0.9\textwidth]{img/Fig1}
  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}

@@

\begin{figure}[ht!]
  \centering
  %\includegraphics[width=0.9\textwidth]{Fig2}
   \includegraphics[width=0.9\textwidth]{img/Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  %\includegraphics[width=0.9\textwidth]{Fig3}
  \includegraphics[width=0.9\textwidth]{img/Fig3}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}
